Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-12-07 Papers/2016/Studia Logica - NaturalDeduction/GodProof-ND.tex
Move accepted Natural Deduction paper

An early unpublished draft of Section \ref{sec:newproof} of this paper,
prepared by Woltzenlogel Paleo in March 2013, served as the starting
point for two distinct but complementary research endeavours. Siders
and Woltzenlogel Paleo transformed that early draft into this paper,
aiming for human-readable and yet formal natural deduction proofs.
Benzm\"uller and Woltzenlogel Paleo, on the other hand, focused on the
verification of variants of G\"odel's ontological arguments using
higher-order automated reasoning tools, which quickly led to many
publications. The main automatically obtained results about G\"odel's
and Scott's proofs appeared in the European Conference on Artificial
Intelligence \citep{ECAI2014} and, two years later, a detailed
analysis of an inconsistency found in G\"odel's axioms was published
in the International Joint Conference on Artificial Intelligence
\citep{IJCAI2016}. The automatic verification of variants by Anderson,
Bjordal and H\'ajek were presented in the First
World Congress on Logic and Religion
\citep{WorldCongressLogicAndReligion} and published in
\citep{controversy} and results related to the modal
collapse were reported in the 4th Word Congress on the Square of
Opposition \citep{SquareOfOpposition} and later published in
\citep{ModalCollapse}. More detailed reconstructions of Scott's
variant within the interactive proof assistants Isabelle and Coq were
released, respectively, in Isabelle's Archive of Formal Proofs
\citep{IsabelleGod} and in the Symposium of Computer Science in Russia
\citep{CSRModalLogicInCoq}. The popularity of the topic led to invited
talks and corresponding invited abstracts in the proceedings of
workshops \citep{WADT,WADTPreProceedings,IWIL} and summer schools
\citep{SpiritualQuest,ReasoningWebSummerSchool} as well as articles
targeting a more general audience \citep{AustrianComputerSociety}.
More recently, this line of research ventured into other types of
ontological arguments, and a computer-assisted reconstruction of an
argument by Leibniz will appear as a chapter in a book dedicated to
the 300th anniversary of Leibniz's death
\citep{LeibnizOntologicalProof}. The automatic verification of G\"odel's
and Scott's variants \citep{ECAI2014} was preceded by an automatic
verification of (a non-modal version of) Anselm's ontological argument
\citep{ZaltaOppenheimer}. However, whereas the complexity of G\"odel's
and Scott's arguments required the embedding of higher-order modal
logics into higher-order logics and the use of higher-order automated
reasoners, Anselm's argument greater simplicity enabled its
verification with a first-order prover. Anselm's argument has also been
formalized within the interactive proof assistant PVS \citep{Rushby}
and there has been a failed attempt to formalize Avicenna's
argument in PVS as well.

@@

Applications of the $\beta$ rule are silently ommitted. Double lines are used to abbreviate propositional reasoning steps when they are tedious and too space-consuming. Dashed lines are used to refer to an axiom or theorem with the proof shown elsewhere. When proof trees are too large to fit on the page, some branches may be displayed separately. In such cases, the conclusion of the branch and the location in the main proof tree where the branch belongs are annotated with the same symbol (a subscripted $\subproof$). Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

@@

In addition to the five axioms and three definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s iteration principle (in the proof of Lemma 2) and the lemma of reflexivity for equality (in the proof of Theorem 1). The use of reflexivity could be avoided, but we chose to use it because it allows the construction of a shorter proof.

@@

Lemma 2 also turns out to be a superfluous detour and therefore the new proof does not include it. Inspecting Scott's proof of Lemma 2, we see that an important step is the derivation of $\nec \ex x. G(x)$ from $\pos \nec \ex x. G(x)$ and {\bf S5}'s iteration principle. Instead, we can derive $\ex x. G(x)$ from $\pos \nec \ex x. G(x)$ and Brouwer's reduction theorem and then derive $\nec \ex x. G(x)$ from $\ex x. G(x)$ by the $\nec_I$ (necessitation) inference rule. This is precisely what the new proof of Theorem 3 does.

@@

Since the iteration principle requires the strong modal logic {\bf S5} while Brouwer's reduction principle requires only the much weaker modal logic {\bf KB},  the possibility of proving Theorem 3 using Brouwer's reduction principle instead of {\bf S5}'s iteration principle is philosophically profound. There have been many discussions of what modal axioms are needed for the ontological argument. In their manuscripts, G\"odel and Scott do not write explicitly which logic they use. However, a proof step in their notes clearly relies on {\bf S5}'s \emph{iteration principle}.
Anderson \citep[footnote 5]{and90} cites several articles questioning the adequacy of {\bf S5} for the ontological argument and conjectures that the ontological argument could be restricted to the weaker modal logic {\bf B}. Sobel \citep[p. 152]{sobel2} acknowledges Anderson's conjecture, claims that it is correct and tries to informally explain how his formal proofs could be modified to rely on {\bf B} instead of {\bf S5}. Nevertheless, neither Anderson nor Sobel presented any formal proof relying only on {\bf B}. The new formal proof presented in this section relies on {\bf KB}, which is weaker than {\bf B}.

@@

Because of Anderson's emendations to Axiom 1 and Definitions 1 and 2, the proof of Lemma 1 has to be modified\footnote{Due to lack of space, the instantiation of Axiom \ref{A:A2} is ommitted in the emended proof of Lemma \ref{A:L1}.} as well.
The proof shown below has been modified as little as possible. Interestingly, in order to eliminate 
the extra box modality in Anderson's emended Definition \ref{A:D1}, the modified proof below uses the modal axiom T. 
Likewise, to obtain the extra box in the emended Definition \ref{A:D2}, Axioms 2 and 3 were used.

@@

%implication for equivalence in essence: 
%\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} \dashedLine
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec  \forall y. (G(y)\imp \varphi(y))$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. (G(y)\imp \varphi(y))$}\RightLabel{$\imp_E$}
\AXC{Axiom \ref{A:A2}} \dashedLine
\UIC{.} \doubleLine \RightLabel{$\all_E$ x 2}
\UIC{.}
\BIC{$P(\varphi)$}
\AXC{$\subproof_{11}$} \dashedLine 
\UIC{$  G(\gamma)$} \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. (P(\varphi)\biimp \nec \varphi(\gamma))$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\imp_I^5$}
\UIC{$[\subproof_{14}] \quad \nec  \forall y. (G(y)\imp \varphi(y))\imp \nec \varphi(\gamma)$}
\end{prooftree}
%\end{small}

@@

\bibitem[{\itshape Anderson \& Gettings (1996)}]{and}
Anderson, C. A.\& Gettings, M. 1996.  {\itshape G\"odel's ontological proof revisited}. In: edited by Hajek P. {\itshape G\"odel '96},  Springer.

@@

\bibitem[{\itshape Benzm\"uller \& Weber \& Woltzenlogel Paleo(2015)}]{WorldCongressLogicAndReligion}
Benzm\"uller, C. \& Weber, L. \& Woltzenlogel Paleo, B. 2015. {\itshape Computer-Assisted Analysis of the Anderson-Hajek Ontological Controversy}. Handbook of the 1st World Congress on Logic and Religion. Eds. Silvestre, R. S. \& B\'{e}ziau, J-Y.

@@

\bibitem[{\itshape Hajek(1996)}]{hajek}
Hajek, P. 1996. Magari and others on G\"odel's Ontological Proof. Eds. Ursini et al. Logic and Logical Algebra, pp. 125--136, Marcel Dekker.

>>>>

An early unpublished draft of Section \ref{sec:newproof} of this paper,
prepared by Woltzenlogel Paleo in March 2013, served as the starting
point for two distinct but complementary research endeavours. Kanckos
and Woltzenlogel Paleo transformed that early draft into this paper,
aiming for human-readable and yet formal natural deduction proofs.
Benzm\"uller and Woltzenlogel Paleo, on the other hand, focused on the
verification of variants of G\"odel's ontological arguments using
higher-order automated reasoning tools, which quickly led to many
publications. The main automatically obtained results about G\"odel's
and Scott's proofs appeared in the European Conference on Artificial
Intelligence \citep{ECAI2014} and, two years later, a detailed
analysis of an inconsistency found in G\"odel's axioms was published
in the International Joint Conference on Artificial Intelligence
\citep{IJCAI2016}. The automatic verification of variants by Anderson,
Bj\o rdal and H\'ajek were presented in the First
World Congress on Logic and Religion
\citep{WorldCongressLogicAndReligion} and published in
\citep{controversy} and results related to the modal
collapse were reported in the 4th Word Congress on the Square of
Opposition \citep{SquareOfOpposition} and later published in
\citep{ModalCollapse}. More detailed reconstructions of Scott's
variant within the interactive proof assistants Isabelle and Coq were
released, respectively, in Isabelle's Archive of Formal Proofs
\citep{IsabelleGod} and in the Symposium of Computer Science in Russia
\citep{CSRModalLogicInCoq}. The popularity of the topic led to invited
talks and corresponding invited abstracts in the proceedings of
workshops \citep{WADT,WADTPreProceedings,IWIL} and summer schools
\citep{SpiritualQuest,ReasoningWebSummerSchool} as well as articles
targeting a more general audience \citep{AustrianComputerSociety}.
More recently, this line of research ventured into other types of
ontological arguments, and a computer-assisted reconstruction of an
argument by Leibniz will appear as a chapter in a book dedicated to
the 300th anniversary of Leibniz's death
\citep{LeibnizOntologicalProof}. The automatic verification of G\"odel's
and Scott's variants \citep{ECAI2014} was preceded by an automatic
verification of (a non-modal version of) Anselm's ontological argument
\citep{ZaltaOppenheimer}. However, whereas the complexity of G\"odel's
and Scott's arguments required the embedding of higher-order modal
logics into higher-order logics and the use of higher-order automated
reasoners, Anselm's argument greater simplicity enabled its
verification with a first-order prover. Anselm's argument has also been
formalized within the interactive proof assistant PVS \citep{Rushby}
and there has been a failed attempt to formalize Avicenna's
argument in PVS as well.

@@

Applications of the $\beta$ rule are silently omitted. Double lines are used to abbreviate propositional reasoning steps when they are tedious and too space-consuming. Dashed lines are used to refer to an axiom or theorem with the proof shown elsewhere. When proof trees are too large to fit on the page, some branches may be displayed separately. In such cases, the conclusion of the branch and the location in the main proof tree where the branch belongs are annotated with the same symbol (a subscripted $\subproof$). Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

@@

In addition to the five axioms and three definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s Iteration principle (in the proof of Lemma 2) and the lemma of reflexivity for equality (in the proof of Theorem 1). The use of reflexivity could be avoided, but we chose to use it because it allows the construction of a shorter proof.

@@

Lemma 2 also turns out to be a superfluous detour and therefore the new proof does not include it. Inspecting Scott's proof of Lemma 2, we see that an important step is the derivation of $\nec \ex x. G(x)$ from $\pos \nec \ex x. G(x)$ and {\bf S5}'s Iteration principle. Instead, we can derive $\ex x. G(x)$ from $\pos \nec \ex x. G(x)$ and Brouwer's reduction theorem and then derive $\nec \ex x. G(x)$ from $\ex x. G(x)$ by the $\nec_I$ (necessitation) inference rule. This is precisely what the new proof of Theorem 3 does.

@@

Since the Iteration principle requires the strong modal logic {\bf S5} while Brouwer's reduction principle requires only the much weaker modal logic {\bf KB},  the possibility of proving Theorem 3 using Brouwer's reduction principle instead of {\bf S5}'s Iteration principle is philosophically profound. There have been many discussions of what modal axioms are needed for the ontological argument. In their manuscripts, G\"odel and Scott do not write explicitly which logic they use. However, a proof step in their notes clearly relies on {\bf S5}'s \emph{Iteration principle}.
Anderson \citep[footnote 5]{and90} cites several articles questioning the adequacy of {\bf S5} for the ontological argument and conjectures that the ontological argument could be restricted to the weaker modal logic {\bf B}. Sobel \citep[p. 152]{sobel2} acknowledges Anderson's conjecture, claims that it is correct and tries to informally explain how his formal proofs could be modified to rely on {\bf B} instead of {\bf S5}. Nevertheless, neither Anderson nor Sobel presented any formal proof relying only on {\bf B}. The new formal proof presented in this section relies on {\bf KB}, which is weaker than {\bf B}.

@@

Because of Anderson's emendations to Axiom 1 and Definitions 1 and 2, the proof of Lemma 1 has to be modified\footnote{Due to lack of space, the instantiation of Axiom \ref{A:A2} is omitted in the emended proof of Lemma \ref{A:L1}.} as well.
The proof shown below has been modified as little as possible. Interestingly, in order to eliminate 
the extra box modality in Anderson's emended Definition \ref{A:D1}, the modified proof below uses the modal axiom T. 
Likewise, to obtain the extra box in the emended Definition \ref{A:D2}, Axioms 2 and 3 were used.

@@

%implication for equivalence in essence: 
%\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} \dashedLine
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec  \forall y. (G(y)\imp \varphi(y))$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. (G(y)\imp \varphi(y))$}
\AXC{Axiom \ref{A:A2}} \dashedLine
\UIC{.} \doubleLine \RightLabel{$\all_E$ x 2}
\UIC{.}\RightLabel{$\imp_E$}
\BIC{$P(\varphi)$}
\AXC{$\subproof_{11}$} \dashedLine 
\UIC{$  G(\gamma)$} \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. (P(\varphi)\biimp \nec \varphi(\gamma))$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\imp_I^5$}
\UIC{$[\subproof_{14}] \quad \nec  \forall y. (G(y)\imp \varphi(y))\imp \nec \varphi(\gamma)$}
\end{prooftree}
%\end{small}

@@

\bibitem[{\itshape Anderson \& Gettings (1996)}]{and}
Anderson, C. A.\& Gettings, M. 1996.  {\itshape G\"odel's ontological proof revisited}. In: edited by H\'ajek P. {\itshape G\"odel '96},  Springer.

@@

\bibitem[{\itshape Benzm\"uller \& Weber \& Woltzenlogel Paleo(2015)}]{WorldCongressLogicAndReligion}
Benzm\"uller, C. \& Weber, L. \& Woltzenlogel Paleo, B. 2015. {\itshape Computer-Assisted Analysis of the Anderson-H\'ajek Ontological Controversy}. Handbook of the 1st World Congress on Logic and Religion. Eds. Silvestre, R. S. \& B\'{e}ziau, J-Y.

@@

\bibitem[{\itshape H\'ajek(1996)}]{hajek}
H\'ajek, P. 1996. Magari and others on G\"odel's Ontological Proof. Eds. Ursini et al. Logic and Logical Algebra, pp. 125--136, Marcel Dekker.
